---
title: "Ph-like ALL supplementary example"
date: "14 oktober 2014"
output:
  html_document:
    theme: flatly
---

## Introduction

This is an R-markdown document. It contains both code and text to help you run examples in R, and create fully reproducible reports!
Here, we're going to read in supplementary data from a recent pediatric-ALL paper in NEJM.

We need the `xlsx` package to read excel files directly. The other option is to export the table you're interested in as a `.csv` file, and read it in directly. 
And the `ggplot2` package for making the plots. 

```{r, echo=FALSE}
# Set your working directory
setwd("~/sandbox/ph-like-rmd-tutorial/")
```

## Loading the data

To load the data, we have two options.

1. Load the excel sheet directly, specifying that we want to read the first sheet. For this we need the `xlsx` package
```{r, eval=FALSE}
# Reading from Excel file
library(xlsx)
table <- read.xlsx("nejmoa1403088_appendix_1.xlsx", 1, na.strings="")
```

2. Load a CSV file or export your excel sheet as csv. In this tutorial, we have already converted the `xlsx` file to a `csv` file to avoid long loading times. Here we're going to load the csv-file from our current working directory.

You can get the data from here: https://ki.box.com/nejm-supplementary

```{r}
# Reading from CSV file with ; as delimiter
data <- read.csv2("Table S1-Tabell 1.csv")
```

***

## Explore your data

First, you may want to know just how much data there actually is in this table: 
```{r}
dim(data)
```

Ok, so we have >2000 cases with >50 different variables recorded.     
What variables are those exactly? 

```{r}
colnames(data)
```

*** 

Now we can start exploring our data, like what's the gender distribution: 
```{r}
table(data$Gender)
barplot(table(data$Gender))
```

And if we want to make more beautiful plots, we can use `ggplot2` instead of base R:
```{r}
library(ggplot2)
ggplot(data, aes(Gender)) + geom_histogram()
```

***

I'm interested in seeing the distribution of different ALL subgroups within their study cohort. 
Also, lets save the table to a variable so we can work on it some more.
```{r}
plot <- ggplot(data, aes(Group)) + geom_bar()
print(plot)
```

The power of ggplot is that we can easily extend our plot, by adding layers.   
For example, its meaningful in ALL to view subgroup distribution by age group. We can easily do that by using `facet_wrap`
```{r}
levels(data$Age)
# Reorder the Age factors so that it goes from young to old, low to high risk
data$Age <- factor(data$Age, levels=c("Childhood SR", "Childhood HR", "Adolescent", "Young adult"))
levels(data$Age)
plot <- ggplot(data, aes(Group, fill=Age)) + geom_bar()
plot <- plot + facet_wrap(~Age)
plot <- plot + theme_minimal()
plot <- plot + theme(axis.text.x = element_text(angle=90, vjust=0.5))
print(plot)
```

This can be extended infinitely
```{r}
data$Kinase.alteration <- factor(data$Kinase.alteration, levels=c("Not tested", "No", "Unknown", "Yes"))
plot <- ggplot(data, aes(Group, fill=Kinase.alteration)) + geom_bar() + facet_wrap(~ Cohort)
print(plot)
summary(data$Kinase.alteration)
```

***

If you'd rather show a pie chart, that is easy

```{r}
pie(table(data$Group))
```

