---
title: "R hands on - handling supplementary material of a paper of interest in R"
author: "Bianca Tesi"
date: "October 1, 2014"
output: html_document
---

**R tutorial** \

1. Download the supplementary material from a paper of interest concerning gene expression data \
The paper that we will use here is **"Chromatin state dynamics during blood formation"** by Lara-Astiago et al. 2014, recently published in Science \

You can download the file (the TableS2 of the supplementary material) from this link: [https://ki.box.com/shared/static/av9p9sbsr7tt9xrdrxu5.txt](https://ki.box.com/shared/static/av9p9sbsr7tt9xrdrxu5.txt) \
If you are curious about the paper, here is the link: [http://www.sciencemag.org/content/345/6199/943.abstract](http://www.sciencemag.org/content/345/6199/943.abstract)

2. Read the file into R, but first:
* install the packages that we will use: *ggplot2* and *reshape2*
* set the working directory to the directory where you have downloaded the file using the setwd command


```{r}
#load libraries
#if you don't have them installed run:
#install.packages("gglot2")
#install.packages("reshape2")

library(ggplot2)
library(reshape2)

setwd("~/Desktop")

data <- read.table("1256271tableS2.txt", header=TRUE, sep="\t")

```


3. Suppose you want to check the expressions of your favorite genes in the cell populations studied in the paper and you want to make a plot \
-
- Have a look at how many rows and column your table has (using the function *dim*) \
- Have a look at the column names to identify the column with the gene name information \
- The see the first 6 lines of the column containing the gene names \
- Then define a vector with your genes of interest  \

```{r}
dim(data)

colnames(data)

head(data$NAME)

mygenes <- c("Vamp1", "Vamp2","Vamp3","Vamp4","Vamp5","Vamp6","Vamp7","Vamp8")
mygenes

```

Now we will only keep the expression data for the genes in your vector "mygenes":
- using the function *which* we can match the rows in the dataframe referring to genes in our vector
- now you can *melt* the created dataset (function within the reshape2 package)
- have a look at it
- as you can see the first 9 rows are not revelant for looking at gene expression, so we cna remove them

```{r}

data_mygenes <- data[which(data$NAME %in% mygenes),]

data_mygenesM <- melt(data_mygenes, id="NAME")
#ignore the warning message

head(data_mygenesM, n=20)

data_mygenesM <- unique(data_mygenesM[10:nrow(data_mygenesM),])

```
Now we are ready to plot.
We will do it using *ggplot2*, an R package that makes nicer plot than the default R package for printing. \

The information you will give to the function *ggplot* are: \
- the name of the dataset \
- the columns that will represent the x and y in your plot \
- the kind of plot (a bar plot) \
- How the different bars should be colored (in this case by cell type) \

```{r}
ggplot(data_mygenesM, aes(data_mygenesM$NAME, as.numeric(data_mygenesM$value))) +    geom_bar(aes(fill = data_mygenesM$variable), position = "dodge", stat="identity")

```
\
You can also make the plot prettier by adding proper axis labels and a title to your plot:
\
```{r}
ggplot(data_mygenesM, aes(data_mygenesM$NAME, as.numeric(data_mygenesM$value))) +    geom_bar(aes(fill = data_mygenesM$variable), position = "dodge", stat="identity") + xlab("") + ylab("Expression values") + ggtitle("Expression of Vamp proteins")

```
